Modeling the Evolution of Mimicry

### 2016-02-29 docs/paper/Paper.tex
Hopefully the final version

According to Langton, Artificial Life is `\textsl{Life made by Man rather than by Nature'}. Taylor also defines it as a tool for biological inquiry \citep{taylor1993}. While providing a brief survey over different AL models he talks about \textsl{Wetware systems} which work at the molecular level, the \textsl{Software systems} which work at the cellular level and the \textsl{Hardware systems} which works at the organism level. The initial contribution of \textsl{software systems} in artificial life was from John von Neumann when he designed the first artificial-life model (without referring to it as such), the famous self-reproducing, computation-universal cellular automata \citep{neumann1966}. He tried to understand the fundamental properties of living systems, especially self-reproduction and the evolution of complex adaptive structures, by constructing simple formal systems that exhibit those properties.

@@

Our model initializes with three kinds of agents. These agents have properties and behavior similar to the \textbf{model}, the \textbf{mimic} and the \textbf{predator}. We represent evolution of pattern for the model and the mimic with the help of \gls{ca}. \gls{ca} can be easily represented by simple rules, which can be expressed as a binary string. The predator will be equipped with a Hopfield network \citep{hopfield1982}, to be able to have pattern recognition capability. The process of evolution will be occurring at the genetic level.

@@

\subsection{FormAL Framework}
The \textsl{``FormAL framework"} is a collection of ideas and concepts taken from Peter Grogono's \gls{formal} project \citep{grogono2003} and are used to build a framework for modeling the evolution of mimicry. In FormAL, an \textbf{Agent} is a simulated organism. It is designed simply, but with capabilities of reproduction using genetic information and modification of genome between generations. There is also interaction between agents while being able to survive and reproduce in a challenging environment.

The framework consists of a three dimensional visual environment where agents get complete freedom of movement defined from their genetic representation. The \textbf{space} of the environment is a three-dimensional lattice of discrete points. The coordinate of each point in space is of the form \((x,y,z) \in \Sigma^3\), where \(\Sigma\) be the set \(\{0, 1, ..., S-1\}\). \(S\) is a universal constant being a small positive integer, which in this simulation has been kept at 20 (the value of the \(World Size\) parameter in table \ref{tab:environment-control-parameters}). \textbf{Time}, being an integer \( (t \geq 0) \), advances in discrete steps in the simulation, where at each step the agents update themselves. The entire three dimensional toroidal environment is divided in multiple cells. A \textbf{cell} is a three dimensional cubical section of the hyperspace. Accordingly in Table \ref{tab:environment-control-parameters} \(ISize\) is the parameter that controls the number of cells in the environment. The size of a cell is evaluated from the entire world size and the total number of cell is calculated as \(ISize^3 = 216\).

\begin{table}[h]
\small
\centering
\begin{tabular}{| l | c |}
	\hline
		\textbf{Parameter} & \textbf{Value} \\ \hline
		ISize: Number of cell in single dimension & 6 \\ \hline
		World Size: Size of a single dimension & 20 \\ \hline
		Cell Size & \( World Size\) / \(ISize \) \\ \hline
		Total Number of Cells & \( ISize^3  = 216\)\\ 
	\hline
\end{tabular}
\caption{Parameters to control the environment.}
\label{tab:environment-control-parameters}
\end{table}

@@

For this simulation the preys are heliconius butterfly and the representation of their wing pattern is with the help of cellular automata (CA). Every prey organism contains a genetic representation of CA with which the predator will identify the prey and store its level of palatability in memory. We choose cellular automata as it can be easily represented with the help of a binary genome and then evolutionary operations on the genomic representation, such as mutation and crossover can easily be applied. Prey's genome is constructed as an 8 bit binary value with a decimal range between 0 to 255. Each of this 256 values have a unique CA pattern associated with it. When storing this pattern in Hopfield memory we take a linear representation of this 2-D pattern, while to consider similarity between two patterns we evaluate the hamming distance between their linear representation.

@@

Predators in the system are designed to provide selection pressure to \textit{models} and \textit{mimics} for the evolution of mimicry. Similar to prey species, they are agents in the FormAL environment capable of mobility and reproduction. In addition, these agents are equipped with Hopfield Network Memory to be able to learn and recognize patterns of the prey species. Their mobility and reproduction capability are controlled at the genetic level, while their memory is not genetically controlled, as it was not possible to come up with a genetic representation for Hopfield Network. Every new predator is born with zero memory and with no inheritance from parents. A set of parameters are defined to control predators' population and learning ability in the environment. Details of which are in table \ref{tab:predator-control-parameters}.

@@

\paragraph{Genome}
Each predator has a 5 bits genome. The first 4 bits represent its mobility and magnitude of the force component with which it moves around the environment. The last bit controls reproduction capability of each species.

\paragraph{Mobility}
Movement behavior of a predator is calculated from its genome. The first 4 bit converted to decimal is the magnitude of force (varying from 0-15) with which it will move towards the maximum crowd of prey present within its neighborhood. If no prey is present in the neighborhood then this force is active in trying to keep predators distributed all over the cells with a constant mobile behavior. This behavior of predator has been designed to increase predator prey interaction in the simulation in terms of one agent chasing the other for survival of species.

@@

@@

% Put the image
\begin{figure*}[t!]
	\centering
	\includegraphics[width=0.77\textwidth]{../tex/images/simTime10k-2Prey}
	\caption[Population distribution of mimicry rings (2 prey species, 10k iterations)]{Population distribution of mimicry rings, initialized with 2 prey species, 10k iterations}
	\label{fig:plot-2-prey}
\end{figure*}

\subsection{Initial configuration with two prey species}
\label{subsec:init-conf-2prey}
% Put the table here
\begin{table}[h]
\small
\centering
\setlength\tabcolsep{2pt}
\begin{tabular}{| p{2cm} | p{1.5cm} | p{1cm} | p{.5cm} | p{1.5cm} | p{.5cm} |}
  \hline
   														&\multicolumn{3}{c|}{Prey configuration} 																	
   														& \multicolumn{2}{c|}{\parbox{2cm}{Predator \\ configuration}} \\ \hline
  \multirow{2}{*}{Population} & Rule110 (Palatable) & \parbox[c]{2.1em}{\includegraphics[scale=0.50]{../tex/images/CARule110}} & 108 
  														& \multicolumn{2}{c|}{\multirow{2}{*}{10}} \\ \cline{2-4}
  					 									& Rule30 (Unpalatable)& \parbox[c]{2.1em}{\includegraphics[scale=0.50]{../tex/images/CARule30}}  & 108 
  					 									& \multicolumn{2}{c|}{}\\ \hline
  \multirow{2}{*}{Reproduction} & Age Limit & \multicolumn{2}{c|}{100}  & \multicolumn{2}{c|}{500} \\ \cline{2-6}
  						 									& Interval  & \multicolumn{2}{c|}{1000} & \multicolumn{2}{c|}{1200} \\ \hline
  \multirow{2}{*}{Mutation Rate} & Pattern   & \multicolumn{2}{c|}{0.05} & \multicolumn{2}{c|}{\multirow{2}{*}{0.3}} \\ \cline{2-4}
  						 									 & Genome    & \multicolumn{2}{c|}{0.5}  & \multicolumn{2}{c|}{} \\ \hline
  Demise Age	 									 & \multicolumn{3}{c|}{2000}							& \multicolumn{2}{c|}{2500} \\ \hline
  Minimum Attack Age						 & \multicolumn{3}{c|}{} 						    & \multicolumn{2}{c|}{500} \\ \hline
  \multirow{2}{*}{\parbox{2cm}{Memory Configuration}} & \multicolumn{3}{c|}{} 					& Minimum & 2 \\ \cline{5-6}
   																			& \multicolumn{3}{c|}{} 					& Maximum & 10 \\ \hline  
\end{tabular}
\caption{Agent configuration of 2 prey species}
\label{tab:config-table-2-prey}
\end{table}

\begin{figure*}[t!]
	\centering
	\includegraphics[width=0.77\textwidth]{../tex/images/simTime8k-4Prey-unp}
	\caption[Population distribution of mimicry rings(4 prey species all unpalatable)]{Population distribution of mimicry rings, initialized with 4 prey species all unpalatable.}
	\label{fig:plot-4-prey-unp}
\end{figure*}

@@

Initial population of predator species has been set to 10 which is in accordance with the prey population in the simulation. The reason for such low number of predator is, unlike prey species which are consumed by predators, there is no cause for the predator species to die accept their natural cause of death, that is to reach their demise age. So predator population can explode very easily. That is why their population is controlled in a restrictive manner with the help of high reproduction age limit and reproduction age interval.

The plot in Figure \ref{fig:plot-2-prey} is simulation time verses prey population after running it for 10000 iterations. With the initial configuration in the above table we can observe that multiple rings of prey population have been created. Two prey species are considered to be in a ring if their CA pattern have Hamming distance within 10 bits. Population of palatable species have been represented with line curve while population of unpalatable species have been presented with dotted curve. Different signs of squares, triangles and diamonds have been used to distinguish between species of prey population. The simulation was initiated with two prey species having CA rule of 30 and 110 and being palatable and unpalatable consecutively. Over time the population of CA Rule 30 dominates the population (Figure \ref{fig:plot-2-prey}) as most predators recognize it as unpalatable. Similarly a palatable population of CA Rule 30 or within the same ring of palatable species starts rising, while at one point overlaps the population of CA Rule 110 (Time: 4000 approx.). CA Rule 110 was initialized as a set of palatable species.

@@

%CARule table
\begin{table}[h]
\small
\centering
\setlength\tabcolsep{2pt}
\begin{tabular}{|l|c|c|c|}
  \hline
  CA Rule & \(60 \equiv 00111100\) & \(61 \equiv 00111101\) & \(62 \equiv 00111110 \) \\ \hline
  Pattern & \parbox[c]{2.1em}{\includegraphics[scale=0.40]{../tex/images/CARule60}} 
  				& \parbox[c]{2.1em}{\includegraphics[scale=0.40]{../tex/images/CARule61}} 
  				& \parbox[c]{2.1em}{\includegraphics[scale=0.40]{../tex/images/CARule62}}\\
  \hline
\end{tabular}
\caption{Difference in prey pattern genotype and phenotype}
\label{tab:diff-in-pattern}
\end{table}

@@

To evaluate the simulation at a more complex level we increased the prey population to 900, consisting of 6 different species with very different pattern configuration. To boost predator-prey interaction we also increased the number of predator population to 30. This resulted in an enormous diversity of species where the total number of mimicry rings reached nearly 50.

@@

In contrast when the simulation was initiated with only palatable species all population of prey have reached its demise at nearly 7000 iterations.

>>>>

According to Langton, Artificial Life is `\textsl{Life made by Man rather than by Nature'}. Taylor also defines it as a tool for biological inquiry \citep{taylor1993}. While providing a brief survey over different AL models he talks about \textsl{Wetware systems} which work at the molecular level, \textsl{Software systems} which work at the cellular level and \textsl{Hardware systems} which works at the organism level. The initial contribution of \textsl{software systems} in artificial life was from John von Neumann when he designed the first artificial-life model (without referring to it as such), the famous self-reproducing, computation-universal cellular automata \citep{neumann1966}. He tried to understand the fundamental properties of living systems, especially self-reproduction and the evolution of complex adaptive structures, by constructing simple formal systems that exhibit those properties.

@@

Our model initializes with three kinds of agents. These agents have properties and behavior similar to the \textbf{model}, the \textbf{mimic} and the \textbf{predator}. We represent evolution of pattern for the model and the mimic with the help of \gls{ca} \citep{Wolfram2002}. \gls{ca} can be easily represented by simple rules, which can be expressed as a binary string. The predator will be equipped with a Hopfield network \citep{hopfield1982}, to have pattern recognition capability. The process of evolution will be occurring at the genetic level.

@@

\subsection{FormAL Framework}
The \textsl{``FormAL framework"} is a collection of concepts taken from Peter Grogono's \gls{formal} project \citep{grogono2003} and are used to build a framework for this model. In FormAL, an \textbf{Agent} is a simulated organism. It is designed simply, but with capabilities of reproduction using genetic information and modification of genome between generations. There is also interaction between agents while being able to survive and reproduce in a challenging environment.

The framework consists of a three dimensional world where agents get complete freedom of movement defined from their genetic representation. This toroidal \textbf{space} is a three-dimensional lattice of discrete points, divided in multiple cells, which can be visualized. A \textbf{cell} is a three dimensional cubical section of the hyperspace. The purpose of the cells is to avoid expensive distance calculations. As two agents are considered \textsl{``close"} to interact when they are in the same cell and \textsl{``distant"} otherwise. \textbf{Time}, being an integer \( (t \geq 0) \), advances in discrete steps in the simulation, where at each step the agents update themselves.

@@

For this simulation the preys are heliconius butterfly and the representation of their wing pattern is with the help of cellular automata (CA). Every prey organism contains a genetic representation of CA with which the predator will identify the prey and store its level of palatability in memory. We choose cellular automata as it can be easily represented with the help of a binary genome and then evolutionary operations on the genomic representation, such as mutation and crossover can easily be applied. Prey's genome is constructed as an 8 bit binary value with a decimal range between 0 to 255. Each of this 256 values have a unique CA pattern associated with it. In generating the pattern of figure \ref{fig:cellular-automata-rule-30}, the genetic representation would be the `New state of center cell'. When storing this pattern in Hopfield memory we take a linear representation of this 2-D pattern, while to consider similarity between two patterns we evaluate the hamming distance between their linear representation.

\begin{figure}[h!]
	\centering
	\includegraphics[scale=1]{images/CARule30}
	\caption[Cellular Automata]{Cellular Automata Rule 30}
	\label{fig:cellular-automata-rule-30}
\end{figure}

\begin{table}[h!]
	\small
	\centering
	\setlength\tabcolsep{2pt}
	\begin{tabular}{| p{3cm} | c | c | c | c | c | c | c | c |}
	  \hline
	  Current Pattern & 111 & 110 & 101 & 100 & 011 & 010 & 001 & 000 \\ \hline
	  New state of center cell & 0 & 0 & 0 & 1 & 1 & 1 & 1 & 0 \\
	  \hline
	\end{tabular}
	\caption{Cellular Automata rule}
	\label{tab:cellular-automata-rule}
\end{table}

@@

Predators in the system are designed to provide selection pressure to \textit{models} and \textit{mimics} for the evolution of mimicry. Similar to prey species, they are agents in the FormAL environment capable of mobility and reproduction. In addition, these agents are equipped with Hopfield Network Memory to be able to learn and recognize patterns of the prey species. Their mobility and reproduction capability are controlled at the genetic level, while their memory is not genetically controlled, as we could not find a suitable encoding for the genetic representation of Hopfield Network. Every new predator is born with zero memory and with no inheritance from parents. A set of parameters are defined to control predators' population and learning ability in the environment (table \ref{tab:predator-control-parameters}).

@@

\paragraph{Genome}
Each predator has a 5 bits genome. The first 4 bits are for mobility, while the last bit controls reproduction capability of each species.

\paragraph{Mobility}
Movement behavior of a predator is calculated from its genome. The first 4 bit converted to decimal is the magnitude of force (varying from 0-15) with which it will move towards the maximum crowd of prey present within its neighborhood. Number of bit are less than prey (6 bits) to reduce their maximum speed. If no prey is present in the neighborhood then this force is active in trying to keep predators distributed all over the cells with a constant mobile behavior. This behavior of predator has been designed to increase predator prey interaction in the simulation in terms of one agent chasing the other for survival of species.

@@

\begin{figure*}[t!]
	\centering
	\includegraphics[width=0.75\textwidth]{images/simTime10k-2Prey}
	\caption[Population distribution of mimicry rings (2 prey species, 10k iterations)]{Population distribution of mimicry rings, initialized with 2 prey species, 10k iterations}
	\label{fig:plot-2-prey}
\end{figure*}

@@

\subsection{Initial configuration with two prey species}
\label{subsec:init-conf-2prey}
% Put the table here
\begin{table}[h]
\small
\centering
\setlength\tabcolsep{2pt}
\begin{tabular}{| p{1.75cm} | m{1.75cm} | p{1cm} | p{.5cm} | p{1.5cm} | p{.5cm} |}
  \hline
   														&\multicolumn{3}{c|}{Prey configuration} 																	
   														& \multicolumn{2}{c|}{\parbox{2cm}{Predator \\ configuration}} \\ \hline
  \multirow{2}{*}{Population} & Rule110 (Palatable) & \parbox[c]{2.1em}{\includegraphics[scale=0.40]{images/CARule110}} & 108 
  														& \multicolumn{2}{c|}{\multirow{2}{*}{10}} \\ \cline{2-4}
  					 									& {\parbox{1.75cm}{Rule30\\ (Unpalatable)}}& \parbox[c]{2.1em}{\includegraphics[scale=0.35]{images/CARule30}}  & 108 
  					 									& \multicolumn{2}{c|}{}\\ \hline
  \multirow{2}{*}{Reproduction} & Age Limit & \multicolumn{2}{c|}{100}  & \multicolumn{2}{c|}{500} \\ \cline{2-6}
  						 									& Interval  & \multicolumn{2}{c|}{1000} & \multicolumn{2}{c|}{1200} \\ \hline
  \multirow{2}{*}{\parbox{1.75cm}{Mutation\\ Rate}} & Pattern   & \multicolumn{2}{c|}{0.05} & \multicolumn{2}{c|}{\multirow{2}{*}{0.3}} \\ \cline{2-4}
  						 									 & Genome    & \multicolumn{2}{c|}{0.5}  & \multicolumn{2}{c|}{} \\ \hline
  Demise Age	 									 & \multicolumn{3}{c|}{2000}							& \multicolumn{2}{c|}{2500} \\ \hline
  {\parbox{1.75cm}{Minimum\\ Attack Age}}	& \multicolumn{3}{c|}{} 						    & \multicolumn{2}{c|}{500} \\ \hline
  \multirow{2}{*}{\parbox{1.75cm}{Memory\\ Configuration}} & \multicolumn{3}{c|}{} 					& Minimum & 2 \\ \cline{5-6}
   																			& \multicolumn{3}{c|}{} 					& Maximum & 10 \\ \hline  
\end{tabular}
\caption{Agent configuration of 2 prey species}
\label{tab:config-table-2-prey}
\end{table}

@@

Initial population of predator species has been set to 10 which is in accordance with the prey population in the simulation. The reason for such low number of predator is, unlike prey species which are consumed by predators, there is no cause for the predator species to die except their natural cause of death, that is to reach their demise age. So predator population can explode very easily. That is why their population is controlled in a restrictive manner with the help of high reproduction age limit and reproduction age interval.

\begin{figure*}[t!]
	\centering
	\includegraphics[width=0.73\textwidth]{images/simTime8k-4Prey-unp}
	\caption[Population distribution of mimicry rings(4 prey species all unpalatable)]{Population distribution of mimicry rings, initialized with 4 prey species all unpalatable.}
	\label{fig:plot-4-prey-unp}
\end{figure*}

The plot in Figure \ref{fig:plot-2-prey} is simulation time verses prey population after running it for 10000 iterations. With the initial configuration in the above table we can observe that multiple rings of prey population have been created. Two prey species are considered to be in a ring if their CA pattern have Hamming distance within 10 bits. Population of palatable species have been represented with line curve while population of unpalatable species have been presented with dotted curve. Different signs of squares, triangles and diamonds have been used to distinguish between species of prey population. The simulation was initiated with two prey species having CA rule of 110 and 30 and being palatable and unpalatable respectively. Over time the population of CA Rule 30 dominates the population (Figure \ref{fig:plot-2-prey}) as most predators recognize it as unpalatable. Similarly a palatable population of CA Rule 30 or within the same ring of palatable species starts rising, while at one point overlaps the population of CA Rule 110 (Time: 4000 approx.). CA Rule 110 was initialized as a set of palatable species.

@@

%CARule table
\begin{table}[h]
\small
\centering
\setlength\tabcolsep{2pt}
\begin{tabular}{|l|c|c|c|}
  \hline
  CA Rule & \(60 \equiv 00111100\) & \(61 \equiv 00111101\) & \(62 \equiv 00111110 \) \\ \hline
  Pattern & \parbox[c]{2.1em}{\includegraphics[scale=0.35]{images/CARule60}} 
  				& \parbox[c]{2.1em}{\includegraphics[scale=0.35]{images/CARule61}} 
  				& \parbox[c]{2.1em}{\includegraphics[scale=0.35]{images/CARule62}}\\
  \hline
\end{tabular}
\caption{Difference in prey pattern genotype and phenotype}
\label{tab:diff-in-pattern}
\end{table}

@@

To evaluate the simulation at a more complex level we increased the prey population to 900, consisting of 6 different species with very different pattern configuration. To boost predator-prey interaction we also increased the number of predator population to 30. This resulted in an enormous diversity of species where the total number of mimicry rings reached nearly 50. Details of this result can be found in \citep{mohiulThesis2011}.

@@

In contrast when the simulation was initiated with only palatable species all population of prey have reached its demise at nearly 7000 iterations, details of which can be found in \citep{mohiulThesis2011}.
